\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-19 NAR_DRAFT.tex
citations in intro

Many methods have been developed for calling somatic mutations from NGS data.
The earliest widely used somatic variant callers developed specifically for tumors, MuTect1 and Varscan2, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants and set a threshold score designed to balance sensitivity and specificity \citep{Koboldt2012,Cibulskis2013}.
Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \citep{Garrison2012},
joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone) \cite{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency-based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE) \citep{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and ensemble and deep learning methods (MutationSeq, SomaticSeq, SNooPer, and BAYSIC) \rngcomment{Citations?}.
These methods vary in their complexity and specific focus.
But they all implicitly or explicitly assume that the rate of mutation is constant across the genome.

The mutational processes that generate single nucleotide variants in tumors do not act uniformly across the genome.
%Single nucleotide variants arise in tumors at a rate and at genomic locations driven by two main processes. 
The processes of spontaneous mutation that are active in all somatic tissues depend sensitively on local nucleotide context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
Additional mutational processes are active in tumors, due to mutagen exposure or defects in DNA maintenance and repair, and these processes are also sensitive to local nucleotide context \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
For example, Figure~\ref{NAR-sigfig}, shows mutation profiles from several tumors.
These real tumor mutation profiles show the diversity of profiles seen even within a single cancer.
%Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.

@@

\subsection{Estimation of the mutation profile}
Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the tri-nucleotide context, which is known to strongly effect the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The tri-nucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot 4$ possible tri-nucleotide contexts $C$.
At the focal site, a mutation $m$ can be to any of three alternate alleles.
Indexing by the $c=\{1 \dots 32\}$ contexts and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
Eq.~\ref{eqn:full_prior} is then
\begin{equation}
  \label{eqn:detailed_prior}
  \begin{array}{rcl}
  \mathrm{p}(\mathrm{S}_{m,c}) &=&  \mathrm{p}(m, M | C = c) \\
                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
  \end{array}
\end{equation}
The first two terms on the right-hand side can be estimated from the observed mutation profile (Figure ~\ref{NAR-sigfig}).
%Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.

We model the observed mutation profile $\mathrm{S}$ (Figure~\ref{NAR-sigfig}\textbf{A}-\textbf{E}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
In a tumor with many high-confidence observed mutations, $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile $S$.
But in practice many entries in $\boldsymbol{\pi}$ would then have zero weight.
We thus model the distribution of $\mathrm{S}$ as Dirichlet-multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$, 
\begin{equation}
\begin{aligned}
  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
  \mathrm{S} \mid \boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
\end{aligned}
\end{equation}
In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely in any context.

@@

\begin{figure}
  \begin{center}
  \includegraphics{figures/sim_signatures_only.pdf}
  \end{center}
  \bkmcomment{tumor in F is probably normal breast plus APOBEC. Mention somewhere?}
  \caption{Tumor mutation profiles.
  In each panel, the x-axis corresponds to each of the 96 possible substitution types, and the y-axis is the proportion of total mutations of each type.
  \textbf{(A)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
  \textbf{(B)} Equal proportions of signatures 1, 4, \& 5.
  \textbf{(C)} Equal proportions of signatures 1, 3, \& 5.
   }
  \label{NAR-simsigfig}
 \end{figure}
\subsection{Tumor simulations}
We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of allele frequencies.
Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{XXX}.
\rngcomment{Put COSMIC website as a citation.}
Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
The ``Concentrated'' profile (Fig.~\ref{NAR-simsigfig}\textbf{A}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
The ``Intermediate'' profile (Fig.~\ref{NAR-simsigfig}\textbf{B}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers \cite{Alexandrov2013}.
The ``Diffuse'' profile (Fig.~\ref{NAR-simsigfig}\textbf{C}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair \cite{Nik-Zainal2016}.
Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
Whole genome (100X depth) and whole exome (500X depth) reads were simulated from the GRCh38 reference genome using VarSim \cite{Mu2015} and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were inserted to create tumors with BAMSurgeon with default parameters~\citep{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:
\begin{scriptsize}
\begin{verbatim}
java -Xmx24g -jar $MUTECT_JAR --analysis_type MuTect \
        --reference_sequence $ref_path \
        --dbsnp $db_snp \
        --enable_extended_output \
        --fraction_contamination 0.00 \
        --tumor_f_pretest 0.00 \
        --initial_tumor_lod -10.00 \
        --required_maximum_alt_allele_mapping_quality_score 1 \
        --input_file:normal $tmp_normal \
        --input_file:tumor $tmp_tumor \
        --out $out_path/$chr.txt \
        --coverage_file $out_path/$chr.cov
\end{verbatim}
\end{scriptsize}
Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.

@@

To test the performance of \batcave, we generated six different tumor/normal pairs, corresponding to 100x whole genomes and 500x whole exomes for three different mutation profiles.
The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{XXX}\textbf{A-C}).
We also tested \batcave using two real cancer data sets, a whole-genome Acute Myeloid Leukemia (AML) \cite{Griffiths2015} and a whole-exome multi-region breast cancer \cite{Shi2018}.
In both, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.

>>>>

Many methods have been developed for calling somatic mutations from NGS data.
The earliest widely used somatic variant callers developed specifically for tumors, MuTect1 and Varscan2, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants and set a threshold score designed to balance sensitivity and specificity \citep{Koboldt2012,Cibulskis2013}.
Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \citep{Garrison2012},
joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone) \cite{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency-based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE) \citep{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and ensemble and deep learning methods (MutationSeq, BAYSIC, SomaticSeq, and SNooPer) \citep{Ding2012,Cantarel2014,Fang2015,Spinella2016}.
These methods vary in their complexity and specific focus.
But they all implicitly or explicitly assume that the rate of mutation is constant across the genome.

The mutational processes that generate single nucleotide variants in tumors do not act uniformly across the genome.
%Single nucleotide variants arise in tumors at a rate and at genomic locations driven by two main processes. 
The processes of spontaneous mutation that are active in all somatic tissues depend sensitively on local nucleotide context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
Additional mutational processes are active in tumors, due to mutagen exposure or defects in DNA maintenance and repair, and these processes are also sensitive to local nucleotide context \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
For example, Figure~\ref{NAR-realsigfig}, shows mutation profiles from several tumors.
These real tumor mutation profiles show the diversity of profiles seen even within a single cancer.
%Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.

@@

\subsection{Estimation of the mutation profile}
Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the tri-nucleotide context, which is known to strongly effect the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The tri-nucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot 4$ possible tri-nucleotide contexts $C$.
At the focal site, a mutation $m$ can be to any of three alternate alleles.
Indexing by the $c=\{1 \dots 32\}$ contexts and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
Eq.~\ref{eqn:full_prior} is then
\begin{equation}
  \label{eqn:detailed_prior}
  \begin{array}{rcl}
  \mathrm{p}(\mathrm{S}_{m,c}) &=&  \mathrm{p}(m, M | C = c) \\
                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
  \end{array}
\end{equation}
The first two terms on the right-hand side can be estimated from the observed mutation profile (Figure ~\ref{NAR-realsigfig}).
%Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.

We model the observed mutation profile $\mathrm{S}$ (Figure~\ref{NAR-realsigfig}\textbf{A}-\textbf{D}) as multinomial with parameter $\boldsymbol{\pi} = \{\pi_{m,c}\}$.
Each element of $\boldsymbol{\pi}$ represents the expected proportion of mutations that are to allele $m$ and in context $c$.
In a tumor with many high-confidence observed mutations, $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile $S$.
But in practice many entries in $\boldsymbol{\pi}$ would then have zero weight.
We thus model the distribution of $\mathrm{S}$ as Dirichlet-multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$, 
\begin{equation}
\begin{aligned}
  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
  \mathrm{S} \mid \boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
\end{aligned}
\end{equation}
In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a priori mutation is equally likely in any context.

@@

\begin{figure}
  \begin{center}
  \includegraphics{figures/sim_signatures_only.pdf}
  \end{center}
  \bkmcomment{tumor in F is probably normal breast plus APOBEC. Mention somewhere?}
  \caption{Tumor mutation profiles.
  In each panel, the x-axis corresponds to each of the 96 possible substitution types, and the y-axis is the proportion of total mutations of each type.
  \textbf{(A)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
  \textbf{(B)} Equal proportions of signatures 1, 4, \& 5.
  \textbf{(C)} Equal proportions of signatures 1, 3, \& 5.
   }
  \label{NAR-simsigfig}
 \end{figure}
\subsection{Tumor simulations}
We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of allele frequencies.
Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) %\cite{XXX}.
\rngcomment{Put COSMIC website as a citation.}
Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
The ``Concentrated'' profile (Fig.~\ref{NAR-simsigfig}\textbf{A}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
The ``Intermediate'' profile (Fig.~\ref{NAR-simsigfig}\textbf{B}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers \cite{Alexandrov2013}.
The ``Diffuse'' profile (Fig.~\ref{NAR-simsigfig}\textbf{C}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair \cite{Nik-Zainal2016}.
Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
Whole genome (100X depth) and whole exome (500X depth) reads were simulated from the GRCh38 reference genome using VarSim \cite{Mu2015} and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were inserted to create tumors with BAMSurgeon with default parameters~\citep{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:
\begin{scriptsize}
\begin{verbatim}
java -Xmx24g -jar $MUTECT_JAR --analysis_type MuTect \
        --reference_sequence $ref_path \
        --dbsnp $db_snp \
        --enable_extended_output \
        --fraction_contamination 0.00 \
        --tumor_f_pretest 0.00 \
        --initial_tumor_lod -10.00 \
        --required_maximum_alt_allele_mapping_quality_score 1 \
        --input_file:normal $tmp_normal \
        --input_file:tumor $tmp_tumor \
        --out $out_path/$chr.txt \
        --coverage_file $out_path/$chr.cov
\end{verbatim}
\end{scriptsize}
Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.

@@

To test the performance of \batcave, we generated six different tumor/normal pairs, corresponding to 100x whole genomes and 500x whole exomes for three different mutation profiles.
The three mutation profiles were chosen to resemble a melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse) (Fig.~\ref{NA-realsigfig}\textbf{A-C}).
We also tested \batcave using two real cancer data sets, a whole-genome Acute Myeloid Leukemia (AML) \citep{Griffith2015} and a whole-exome multi-region breast cancer \cite{Shi2018}.
In both, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
